---
title: "Midterm 2"
author: "Jala Atufa"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your code should be organized, clean, and run free from errors. Be sure to **add your name** to the author header above. You may use any resources to answer these questions (including each other), but you may not post questions to Open Stacks or external help sites. There are 10 total questions.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean! Your plots should use consistent aesthetics throughout. Feel free to be creative- there are many possible solutions to these questions!  

This exam is due by **12:00p on Tuesday, February 23**.  

## Load the libraries
```{r}
library(tidyverse)
library(janitor)
library(here)
options(scipen=999) #disables scientific notation when printing
```
```{r}
library(RColorBrewer)
library(paletteer)
```


## Gapminder
For this assignment, we are going to use data from  [gapminder](https://www.gapminder.org/). Gapminder includes information about economics, population, social issues, and life expectancy from countries all over the world. We will use three data sets, so please load all three.  

One thing to note is that the data include years beyond 2021. These are projections based on modeling done by the gapminder organization. Start by importing the data.
```{r}
population <- read_csv(here("midterm2", "data", "population_total.csv"))
population
```

```{r}
income <- read_csv(here("midterm2", "data", "income_per_person_gdppercapita_ppp_inflation_adjusted.csv"))
income
```

```{r}
life_expectancy <- read_csv(here("midterm2", "data", "life_expectancy_years.csv"))
life_expectancy
```

1. (3 points) Once you have an idea of the structure of the data, please make each data set tidy and store them as new objects. You will need both the original and tidy data!
```{r}
income_year<-income %>% 
  pivot_longer(-country,
               names_to = "year",
               values_to="income")
income_year
```

```{r}
life_expectancy_year<-life_expectancy %>% 
  pivot_longer(-country,
               names_to = "year",
               values_to="life_expectancy")
life_expectancy_year
```

```{r}
population_year <- population %>% 
  pivot_longer(-country,
               names_to="year",
               values_to="population")
population_year
```

2. (1 point) How many different countries are represented in the data? Provide the total number and their names. Since each data set includes different numbers of countries, you will need to do this for each one.
```{r}
income %>% 
  count() %>% 
  glimpse()
```

```{r}
life_expectancy %>% 
  count()
```

```{r}
population %>% 
  count()
```

## Life Expectancy  

3. (2 points) Let's limit the data to the past 100 years (1920-2020). For these years, which country has the highest life expectancy? How about the lowest life expectancy?  
For this question, I decided to examine the mean max and min life expectancies by taking the averages over the last 100 years. 
Lowest
```{r}
life_expectancy_year %>% 
  filter(between(year, 1920,2020)) %>%
  group_by(country) %>% 
  summarise(mean_life_exp = mean(life_expectancy)) %>% 
  top_n(n=-5)
```

Highest
```{r}
life_expectancy_year %>% 
  filter(between(year, 1920,2020)) %>%
  group_by(country) %>% 
  summarise(mean_life_exp = mean(life_expectancy)) %>% 
  top_n(n=5)
```

4. (3 points) Although we can see which country has the highest life expectancy for the past 100 years, we don't know which countries have changed the most. What are the top 5 countries that have experienced the biggest improvement in life expectancy between 1920-2020?
For this question, I decided to look specifically at only the years 1920 and 2020 since this is the beginning and end of the period we are looking at. While I know other things happened inbetween this time period, those years are the ones I looked at. 
```{r}
life_expectancy_year %>% 
  filter(year == 1920 |year == 2020) %>% 
  group_by(country) %>% 
  summarise(min_life_exp = min(life_expectancy, na.rm=T),
            max_life_exp = max(life_expectancy, na.rm=T),
            change_life_exp = (max_life_exp-min_life_exp)) %>% 
  top_n(5, change_life_exp)
```
If we look at 1920 to 2020, Kuwait had the biggest change in life expectancy. 

5. (3 points) Make a plot that shows the change over the past 100 years for the country with the biggest improvement in life expectancy. Be sure to add appropriate aesthetics to make the plot clean and clear. Once you have made the plot, do a little internet searching and see if you can discover what historical event may have contributed to this remarkable change.  

```{r}
life_expectancy_year %>% 
  filter(country == "Kuwait", between(year, 1920, 2020)) %>% 
  ggplot(aes(x= year, y= life_expectancy, color = country))+
  geom_point(na.rm=T)+
  labs(title = "Kuwait Life Expectancy from 1920 to 2020",
       x = "Year",
       y = "Life Expectancy")+
  theme_linedraw()+
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text.x = element_text(angle = 70, hjust = 1))+
  scale_x_discrete(breaks=c("1920", "1925", "1930", "1935", "1940", "1945", "1950", "1955", "1960", "1965", "1970", "1975", "1980", "1985", "1990", "1995", "2000", "2005", "2010", "2015", "2020"))
```
Kuwait has had a serious improvement in life expectancy as trade has boomed as oil, a commodity they have plenty of, has become more and more essential in their every day lives. 

## Population Growth
6. (3 points) Which 5 countries have had the highest population growth over the past 100 years (1920-2020)?
```{r}
population_year %>% 
   filter(year == 1920 |year == 2020) %>% 
  group_by(country) %>% 
  summarise(min_pop = min(population, na.rm=T),
            max_pop = max(population, na.rm=T),
            change_pop = (max_pop-min_pop)) %>% 
  top_n(5, change_pop)
```

7. (4 points) Produce a plot that shows the 5 countries that have had the highest population growth over the past 100 years (1920-2020). Which countries appear to have had exponential growth?  
```{r}
population_year %>% 
  filter(country == "China" |country == "India" | country == "Indonesia" | country == "Pakistan" | country == "United States", between(year, 1920, 2020)) %>% 
  ggplot(aes(x= year, y= population, color= country))+
  geom_point(na.rm=T)+
  labs(title = "Greatest Population Growth from 1920 to 2020",
       x = "Year",
       y = "Population")+
  theme_linedraw()+
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text.x = element_text(angle = 70, hjust = 1))+
  scale_x_discrete(breaks=c("1920", "1925", "1930", "1935", "1940", "1945", "1950", "1955", "1960", "1965", "1970", "1975", "1980", "1985", "1990", "1995", "2000", "2005", "2010", "2015", "2020"))
```
China and India appeared to have an exponential growth and significant increase in population size. 

## Income
The units used for income are gross domestic product per person adjusted for differences in purchasing power in international dollars.

8. (4 points) As in the previous questions, which countries have experienced the biggest growth in per person GDP. Show this as a table and then plot the changes for the top 5 countries. With a bit of research, you should be able to explain the dramatic downturns of the wealthiest economies that occurred during the 1980's.
Looking at the beginning of the timeframe compared to the end of the timeframe you see - 
```{r}
income_gdp <-income_year %>% 
  group_by(country) %>% 
  filter(year == 1920 | year ==2020)%>%
  summarise(min_inc = min(income, na.rm=T),
            max_inc = max(income, na.rm=T),
            change_inc = (max_inc-min_inc)) %>% 
  top_n(5, change_inc)
income_gdp  
```
```{r}
income_year %>% 
  filter(country == "Qatar" | country == "Luxembourg" | country == "Singapore" | country == "Brunei" | country == "Ireland" , between(year, 1920, 2020)) %>%
  ggplot(aes(x = year, y = income, group = country, color = country))+
  geom_point(na.rm=T)+
  labs(title = "Greatest Income Growth from 1920 to 2020",
       x = "Year",
       y = "Income")+
  theme_linedraw()+
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        axis.text.x = element_text(angle = 70, hjust = 1))+
  scale_x_discrete(breaks=c("1920", "1925", "1930", "1935", "1940", "1945", "1950", "1955", "1960", "1965", "1970", "1975", "1980", "1985", "1990", "1995", "2000", "2005", "2010", "2015", "2020"))
  
```
Qatar actually gained it's independence from Britain in the 1970's - around the time of its huge growth in income - it's fall in income may be due to recession in the time or perhaps wars in close-proximity to the country. 

9. (3 points) Create three new objects that restrict each data set (life expectancy, population, income) to the years 1920-2020. Hint: I suggest doing this with the long form of your data. Once this is done, merge all three data sets using the code I provide below. You may need to adjust the code depending on how you have named your objects. I called mine `life_expectancy_100`, `population_100`, and `income_100`. For some of you, learning these `joins` will be important for your project.  

life_expectancy_100
```{r}
life_expectancy_100 <- life_expectancy_year %>%
  filter(between(year, 1920, 2020))
life_expectancy_100
```

population_100
```{r}
population_100 <- population_year %>%
  filter(between(year, 1920, 2020))
population_100
```

income_100
```{r}
income_100 <- income_year %>%
  filter(between(year, 1920, 2020))
income_100
```

```{r}
gapminder_join <- inner_join(life_expectancy_100, population_100, by= c("country", "year"))
gapminder_join <- inner_join(gapminder_join, income_100, by= c("country", "year"))
gapminder_join
```

10. (4 points) Use the joined data to perform an analysis of your choice. The analysis should include a comparison between two or more of the variables `life_expectancy`, `population`, or `income.`
For my analysis I wanted to look at correlations of income and population and decided to focus on how it related to the most recent year - 2020. We can see on this graph that there is a positive correlation between what the income in a country is and what size population it can support but we also see many countries with low incomes and high populations. 
```{r}
glimpse(gapminder_join)
```
```{r}
gapminder_join %>% 
  filter(year == 2020) %>% 
  group_by(country) %>% 
  ggplot(aes(x = income, y=life_expectancy))+
  geom_point(na.rm=T)+
  labs(title = "Income v. Population in 2020",
       x = "Income",
       y = "Population")+
  theme_linedraw()+
  theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10))
```

